---
title: "Install R package ctrdata"
author: "Ralf Herold"
date: 2021-05-08
output:
  rmarkdown::pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
  rmarkdown::html_vignette: 
    toc: true
vignette: >
  %\VignetteIndexEntry{Install R package ctrdata}
  %\VignetteKeywords{ctrdata,vignette}
  %\VignettePackage{ctrdata}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#
knitr::opts_chunk$set(
  eval = FALSE
  )
#
```

# Install package `ctrdata` on a R system

The R Project website ([https://www.r-project.org/](https://www.r-project.org/)) provides installers for the R system. It can be used with software products and graphical user interfaces such as [R Studio](https://www.rstudio.com/products/RStudio/), or [Microsoft R Open](https://mran.microsoft.com/open/), or from [Visual Studio Code](https://marketplace.visualstudio.com/search?term=R&target=VSCode&category=All%20categories&sortBy=Relevance). 

General information on the `ctrdata` package is available here: [https://github.com/rfhb/ctrdata](https://github.com/rfhb/ctrdata). 

In R, execute: 

```{r install_ctrdata, eval=FALSE}
install.packages("ctrdata")
```

For using the development version of package `ctrdata`, install from GitHub: 

```{r, eval=FALSE}
# install preparatory package
install.packages(c("devtools", "httr"))
devtools::install_github("rfhb/ctrdata")
```

Either of the above should install package `ctrdata` into the user's library. 

# Internet access via proxy

Functions in package `ctrdata` that start with `ctr...` require access to trial registers over the internet via the `https` protocol. Package `ctrdata` checks and _automatically uses_ the proxy that is set under MS Windows in system settings. However, proxy settings need to be set by the user for other operating systems and for more complicated environments such as using authenticating proxies, as follows: 

```{r, eval=FALSE}
Sys.setenv(https_proxy = "your_proxy.server.domain:8080")
Sys.setenv(https_proxy_user = "userid:password") # if needed
```

# Additional installation aspects for MS Windows

On MS Windows, it seems recommended to not use UNC notation (such as `\\server\directory`) for specifying the user's library location, but the following notation using a local disk or mapped network share. If the user's library location needs to be set, use: 

```{r, eval=FALSE}
.libPaths("D:/my/directory/")
```

Package `ctrdata` requires on MS Windows that the cygwin environment be installed, into the local directory `c:\cygwin` (or any folder corresponding to `c:\cygw*`). The applications php, bash, perl, cat and sed in the cygwin environment are required for the function `ctrLoadQueryIntoDb()` of package `ctrdata`. No other function in this package has this requirement. The installation of a minimal cygwin environment on MS Windows can be done with the help of package `ctrdata` in R as follows: 

```{r, eval=FALSE}
ctrdata::installCygwinWindowsDoInstall() 
```

If internet access has to be via a proxy and this was not correctly detected automatically, it can be specified for the installation of cygwin as follows: 

```{r, eval=FALSE}
ctrdata::installCygwinWindowsDoInstall(proxy = "proxy.server.domain:8080") 
```

Users who want or need to install cygwin manually, can download the setup executable from [here](https://cygwin.com/install.html). In a MS Windows command window or Powershell window, use the following command line. The parameters are explained [here](https://cygwin.com/faq/faq.html#faq.setup.cli). 

```{bash, eval=FALSE}
setup-x86_64.exe --no-admin --quiet-mode --verbose --upgrade-also --root c:/cygwin 
--site http://www.mirrorservice.org/sites/sourceware.org/pub/cygwin/ 
--packages perl,php-jsonc,php-simplexml
```

In typical Linux and macOS environments, the required tools php, bash, perl, cat and sed are installed and available by default. 

# Database

At this time, an SQLite or a remote or a local MongoDB database can be used with the package `ctrdata`. SQLite is provided in the R package [`RSQLite`](https://rsqlite.r-dbi.org/). Suggested installation instructions for a local MongoDB server are [here](https://docs.mongodb.com/manual/administration/install-community/); a remote MongoDB database server is accessible [here](https://www.mongodb.com/cloud/atlas). 

# Attach package `ctrdata`

After completing the installation, attach the package from the user's library. This will also check the availability of the additional installation requirements as mentioned above: 

```{r attach_ctrdata}
library(ctrdata)
```

Remember to respect the registers' terms and conditions (see `ctrOpenSearchPagesInBrowser(copyright = TRUE)`). Please cite this package in any publication as follows: Ralf Herold (2020). ctrdata: Retrieve and Analyze Clinical Trials in Public Registers. R package version 1.4, https://cran.r-project.org/package=ctrdata
