---
title: "ctrdata getting started"
author: "Ralf Herold"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Getting started with R package `ctrdata` for clinical trial protocol-related information

```{r, include=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```

* Install package `ctrdata` from a vanilla system: 
```{r}
install.packages(c("devtools", "httr"))
#
# set proxy if needed
library(httr)
set_config(use_proxy("proxy.server.domain", 8080))
#
# on windows change library path to not use an UNC notation (\\server\directory)
.libPaths("D:/my/directory/")
#
devtools::install_github("rfhb/ctrdata")
#
```


* Attach package `ctrdata`: 
```{r}
library(ctrdata)
```

* Open register's advanced search page in browser: 
```{r}
#
ctrOpenSearchPagesInBrowser()
#
# Please review and respect register copyrights:
#
ctrOpenSearchPagesInBrowser(copyright = TRUE)
#
```

* Click search parameters and execute search in browser 

* Copy address from browser address bar to clipboard

* Get address from clipboard: 
```{r}
#
q <- ctrGetQueryUrlFromBrowser()
#
# Found search query from EUCTR.
# [1] "cancer&age=under-18"
#
```

* Retrieve protocol-related information, transform, save to database and analyse:
```{r}
#
ctrLoadQueryIntoDb(q)
#
# If no parameters are given for a database connection: uses mongodb
# on localhost, port 27017, database "users", collection "ctrdata"
# note: when run for first time, may download variety.js
#
# Show which queries have been downloaded into the database so far
#
ctrQueryHistoryInDb()
#
# Total number of records: 6143
# Number of queries in history: 2
#       query-timestamp query-register query-records                  query-term
# 1 2016-01-13-10-51-56          CTGOV          5233 type=Intr&cond=cancer&age=0
# 2 2016-01-13-10-40-16          EUCTR           910         cancer&age=under-18
#
# find names of fields of interest in database:
#
dbFindVariable("date")
#
# Returning first of 20 keys found.
# [1] "firstreceived_date"
#
# Get all records that have values in all specified fields.
# Note that b31_... is an element within the array b1_...
#
result <- dbGetVariablesIntoDf(c("b1_sponsor.b31_and_b32_status_of_the_sponsor", "x5_trial_status"))
#
# Tabulate the status of the clinical trial on the date of information retrieval
#
with (result, table (x5_trial_status, b31_and_b32_status_of_the_sponsor))
#
#                     b31_and_b32_status_of_the_sponsor
# x5_trial_status      Commercial Non-Commercial
#   Completed                 138             30
#   Not Authorised              3              0
#   Ongoing                   339            290
#   Prematurely Ended          35              4
#   Restarted                   8              0
#   Temporarily Halted         14              4
#
```

