---
title: "Install R package ctrdata"
author: "Ralf Herold"
date: 2022-03-20
output:
  rmarkdown::pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Install R package ctrdata}
  %\VignetteKeywords{ctrdata,vignette}
  %\VignettePackage{ctrdata}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Install package `ctrdata` on a R system

The R Project website ([https://www.r-project.org/](https://www.r-project.org/)) provides installers for the R system. It can be used with software products and graphical user interfaces such as [R Studio](https://www.rstudio.com/products/RStudio/), or [Microsoft R Open](https://mran.microsoft.com/open/), or from [Visual Studio Code](https://marketplace.visualstudio.com/items?itemName=Ikuyadeu.r).

General information on the `ctrdata` package is available here: [https://github.com/rfhb/ctrdata](https://github.com/rfhb/ctrdata).

In R, execute:

```{r install_ctrdata, eval=FALSE}
install.packages("ctrdata")
```

For using the development version of package `ctrdata`, install from GitHub:

```{r, eval=FALSE}
# install preparatory package
install.packages(c("devtools", "httr"))
devtools::install_github("rfhb/ctrdata")
```

Either of the above should install package `ctrdata` into the user's library.

## Internet access via proxy

Functions in package `ctrdata` that start with `ctr...` require access to trial registers over the internet via the `https` protocol. Many organisastions use transparent proxies that may _not_ require users to do any configurations. However, if necessary, package `ctrdata` can use proxy settings set by users in their R session such as follows:

```{r, eval=FALSE}
Sys.setenv(https_proxy = "your_proxy.server.domain:8080") # if needed
Sys.setenv(https_proxy_user = "your_userid:your_password") # if needed
```

## Additional installation aspects

The command line tools `perl`, `sed` and `php` (5.2 or higher) are required for `ctrLoadQueryIntoDb()`, the main function of package `ctrdata`. No other function in this package has this requirement.

### MS Windows

For this requirement on MS Windows, the [`Cygwin`](https://cygwin.org/install.html) environment has to be installed, into the local directory `c:\cygwin` (or any folder corresponding to `c:\cygw*`). The installation of a minimal Cygwin environment on MS Windows can be done with the help of package `ctrdata` in R as follows:

```{r, eval=FALSE}
ctrdata::installCygwinWindowsDoInstall()
```

To update or correct the Cygwin installation:

```{r, eval=FALSE}
ctrdata::installCygwinWindowsDoInstall(force = TRUE)
```

If internet access requires to specify a proxy (see above):

```{r, eval=FALSE}
ctrdata::installCygwinWindowsDoInstall(proxy = "proxy.server.domain:8080")
```

To install Cygwin manually, download the setup executable from [here](https://cygwin.com/install.html). In a MS Windows command window or Powershell window, use the following command line. The parameters are explained [here](https://cygwin.com/faq/faq.html#faq.setup.cli).

```{bash, eval=FALSE}
setup-x86_64.exe --no-admin --quiet-mode --verbose --upgrade-also --root c:/cygwin
--site http://www.mirrorservice.org/sites/sourceware.org/pub/cygwin/
--packages perl,php-jsonc,php-simplexml
```

### macOS, Linux and other operating systems

The command line tools `perl`, `sed` and `php` (5.2 or higher) may already be available by default in some Linux and macOS systems. They are checked when running `ctrLoadQueryIntoDb()`; alternatively, check with `(ctrdata:::checkBinary())`.  

- In macOS, [`homebrew`](https://brew.sh/) can be used to install `php` which may be needed starting from macOS 12 Monterey: first install [`homebrew`](https://brew.sh/) and then run in Terminal the command: `brew install php`.

- In Linux, tools for installation vary by distribution (e.g., `sudo apt install php php-xml php-json`)

## Databases to use

At this time, a PostgreSQL, an SQLite or a remote or local MongoDB database can be used with the package `ctrdata`. A full SQLite database is provided in the R package [`RSQLite`](https://rsqlite.r-dbi.org/). Suggested installation instructions for PostgreSQL are [here](https://www.postgresql.org/download/) and for a local MongoDB server are [here](https://docs.mongodb.com/manual/administration/install-community/); a remote MongoDB database server is accessible [here](https://www.mongodb.com/cloud/atlas). 

If available, a PostgreSQL or MongoDB local server is recommended for notably faster performance than SQLite. 

Purpose | Function call | Reference
-------------------- | -------------------- | ----------------
Create SQLite database connection | `dbc <- nodbi::src_sqlite(dbname = "name_of_my_database", collection = "name_of_my_collection")` | `nodbi::src_sqlite()`
Create PostgreSQL database connection | `dbc <- nodbi::src_postgres(dbname = "name_of_my_database")`; `dbc[["collection"]] <- "name_of_my_collection"`  | `nodbi::src_postgres()`
Create MongoDB database connection | `dbc <- nodbi::src_mongo(db = "name_of_my_database", collection = "name_of_my_collection")`  | `nodbi::src_mongo()`
Subsequently, use connection `dbc` with `ctrdata` functions | `ctrdata::{ctr,db}*(con = dbc, ...)` |

## Attach package `ctrdata`

After completing the installation, attach the package from the user's library. This will also check the availability of the additional installation requirements as mentioned above:

```{r attach_ctrdata}
library(ctrdata)
```

Remember to respect the registers' terms and conditions (see `ctrOpenSearchPagesInBrowser(copyright = TRUE)`). Please cite this package in any publication as follows: Ralf Herold (2022). ctrdata: Retrieve and Analyze Clinical Trials in Public Registers. R package version 1.9.0. https://cran.r-project.org/package=ctrdata
