---
title: "Retrieve clinical trial information"
author: "Ralf Herold"
date: 2023-06-24
output:
  rmarkdown::pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Retrieve clinical trial information}
  %\VignetteKeywords{ctrdata,vignette}
  %\VignettePackage{ctrdata}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Get started

### Attach package `ctrdata`

```{r attach_ctrdata}
library(ctrdata)
citation("ctrdata")
```

Remember to respect the registers' terms and conditions (see `ctrOpenSearchPagesInBrowser(copyright = TRUE)`). Please cite this package in any publication as follows: Ralf Herold (2022). ctrdata: Retrieve and Analyze Clinical Trials in Public Registers. R package version 1.9.0. https://cran.r-project.org/package=ctrdata

### Open register's advanced search page in browser

These functions open the browser, where the user can start searching for trials of interest.

```{r show_brower_search_pages}
# Please review and respect register copyrights:
ctrOpenSearchPagesInBrowser(
  copyright = TRUE
)
# Open browser with example search:
ctrOpenSearchPagesInBrowser(
  url = "cancer&age=under-18",
  register = "EUCTR"
)
```

### Adjust search parameters and execute search in browser

Refine the search until the trials of interest are listed in the browser. The total number of trials that can be retrieved with package `ctrdata` is intentionally limited to queries with at most 10000 result records.

### Copy address from browser address bar to clipboard

Use functions or keyboard shortcuts according to the operating system. See [here](https://rfhb.github.io/ctrdata/#id_3-script-to-automatically-copy-users-query-from-web-browser) for our automation to copy the URLs of a user's queries in any of the supported clinical trial registers. 

### Get address from clipboard

The next steps are executed in the R environment:

```{r, include=FALSE}
q <- "https://www.clinicaltrialsregister.eu/ctr-search/search?query=cancer&age=under-18&status=completed&phase=phase-one"
clipr::write_clip(q)
```


```{r get_query_from_browser}
q <- ctrGetQueryUrl()
# * Using clipboard content as register query URL: https://www.clinicaltrialsregister.eu/ctr-search/search?query=cancer&age=under-18&status=completed&phase=phase-one
# * Found search query from EUCTR: query=cancer&age=under-18&status=completed&phase=phase-one

q
#                                                   query-term  query-register
# 1 query=cancer&age=under-18&status=completed&phase=phase-one           EUCTR

# To check, this opens a browser with the query
ctrOpenSearchPagesInBrowser(url = q)
```

### Retrieve protocol-related information, transform, save to database, check

```{r execute_load_query, eval=FALSE}
# Count number of trial records
ctrLoadQueryIntoDb(
  queryterm = q,
  only.count = TRUE
)$n
# (1/3) Checking trials in EUCTR:
# Retrieved overview, multiple records of 88 trial(s) from 4 page(s) to be downloaded
# [1] 88

# Connect to a database and chose a table / collection
db <- nodbi::src_sqlite(
  dbname = "sqlite_file.sql",
  collection = "test"
)

# Retrieve records, download into database
ctrLoadQueryIntoDb(
  queryterm = q,
  con = db
)
# Checking helper binaries: done
# (1/3) Checking trials in EUCTR:
# Retrieved overview, multiple records of 88 trial(s) from 5 page(s) to be downloaded (estimate: 4.4 MB)
# Downloading trials...
# Note: register server cannot compress data, transfer takes longer, about 0.4s per trial
# Download status: 5 done; 0 in progress. Total size: 6.99 Mb (100%)... done!             
# (2/3) Converting to JSON, 339 records converted
# (3/3) Importing JSON records into database...
# = Imported or updated 339 records on 88 trial(s) 
# No history found in expected format.
# Updated history ("meta-info" in "test")

# Show which queries have been downloaded into database
dbQueryHistory(con = db)
#       query-timestamp query-register query-records
# 1 2023-05-07 09:41:36          EUCTR           339
#                                                   query-term
# 1 query=cancer&age=under-18&status=completed&phase=phase-one
```

With a file-base SQLite database, this takes about 20 seconds for about 300 records, with most of the time needed for internet-retrieval with is slow from this register. Speed is higher with other registers, with using MongoDB and with memory-based SQLite.

## Repeat and update a previous query

```{r}
ctrLoadQueryIntoDb(
  querytoupdate = "last",
  con = db
)
```

Instead of "last", an integer number can be specified for `querytoupdate` that corresponds to the number when using `dbQueryHistory()`.

Depending on the register, an update (differential update) is possible or the original query is executed fully again.

## Retrieve results

For EUCTR, result-related trial information has to be requested to be retrieved, because it will take longer to download and store. For CTGOV, any results are always included in the retrieval.

```{r}
ctrLoadQueryIntoDb(
  queryterm = q,
  euctrresults = TRUE,
  con = db
)
# Checking helper binaries: done
# (1/3) Checking trials in EUCTR:
# Retrieved overview, multiple records of 88 trial(s) from 5 page(s) to be downloaded (estimate: 4.4 MB)
# Downloading trials...
# Note: register server cannot compress data, transfer takes longer, about 0.3s per trial
# Download status: 5 done; 0 in progress. Total size: 6.99 Mb (100%)... done!             
# (2/3) Converting to JSON, 339 records converted
# (3/3) Importing JSON records into database...
# = Imported or updated 339 records on 88 trial(s) 
# Retrieving results if available from EUCTR for 88 trials: 
# (1/4) Downloading and extracting results (. = data, F = file[s] and data, x = none):
# Download status: 88 done; 0 in progress. Total size: 55.07 Mb (100%)... done!             
# F . F F . F . F . F . . . F . . . . . . F . . . . . . . . . . . . . . . . . . . 
# . . . . . . . . F . . . . . F . . F . . . . . . F . . . . . 
# (2/3) Converting to JSON, 70 records converted
# (3/4) Importing JSON into database...
# (4/4) Results history: not retrieved (euctrresultshistory = FALSE)
# = Imported or updated results for 70 trials
# Updated history ("meta-info" in "test")
```

With a file-base SQLite database, this takes about 4 minutes for about 300 records, with most of the time needed for merging result- and protocol-related information in SQLite; this is much faster with MongoDB and PostgreSQL.

The download or presence of results is not recorded in `dbQueryHistory()` because the availability of results increases over time.

## Add information from another register

The same collection can be used to store (and analyse) trial information from different registers. At the moment, `ctrdata` supports three registers: https://ClinicalTrials.Gov/, https://ClinicalTrialsRegister.EU/,  https://www.ISRCTN.com/ and https://euclinicaltrials.eu/. Example:

```{r}
ctrLoadQueryIntoDb(
  queryterm = "https://classic.clinicaltrials.gov/ct2/results?cond=neuroblastoma&recrs=e&age=0&intr=Drug",
  con = db
)
# * Found search query from CTGOV: cond=neuroblastoma&recrs=e&age=0&intr=Drug
# Checking helper binaries: done
# (1/3) Checking trials in CTGOV:
# Retrieved overview, records of 207 trial(s) are to be downloaded (estimate: 1.7 MB)
# Download status: 1 done; 0 in progress. Total size: 1.43 Mb (100%)... done!             
# (2/3) Converting to JSON, 207 records converted
# (3/3) Importing JSON records into database...
# = Imported or updated 182 trial(s)               
# Updated history ("meta-info" in "test")

# Warning message:
# [...]/ctrDATA15bc13c33351a/ctgov_trials_7.ndjson: Error : C stack usage  8336652 is too close to the limit
```

With a file-base SQLite database, this takes about 10 seconds for about 200 records.

Note that in this example, a warning message is issued from importing an NDJSON file with trial records. The warning arises from the high level of complexity of some of the XML content of some of the trial records. The issue can be resolved by increasing in the operating system the stack size available to R, see: https://github.com/rfhb/ctrdata/issues/22


## Add records from CTIS register into the same collection

As of April 2023, more than 160 trials are publicly accessible in CTIS, which has to be used since January 2023 for new clinical trials in the EU. Queries in the CTIS search interface can be automatically copied to the clipboard so that a user can paste them into `queryterm`, see [here](https://rfhb.github.io/ctrdata/#id_3-script-to-automatically-copy-users-query-from-web-browser). 

```{r}
# Retrieve trials from another register:
ctrLoadQueryIntoDb(
  queryterm = "ageGroupCode=3",
  register = "CTIS",
  con = db
)
# * Found search query from CTIS: ageGroupCode=3, found 141 trials
# (2/5) Downloading and processing part I and parts II... (estimate: 21.15 Mb)
# Download status: 141 done; 0 in progress. Total size: 20.88 Mb (100%)... done!             
# . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . 
# . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . 
# . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . 
# . . . . . . . . . . . . . . . . . . . . . . . . 
# (3/5) Downloading and processing additional data: 
# - publicevents
# - summary
# - layperson
# - csr
# - cm
# - inspections
# - publicevaluation 
# Download status: 221 done; 0 in progress. Total size: 6.45 Mb (100%)... done!             
# 141
# (4/5) Importing JSON records into database...
# (5/5) Updating with additional data: . . .       
# = Imported / updated 141 / 141 / 2 / 141 records on 141 trial(s)
# Updated history ("meta-info" in "test")
```


## Add personal annotations

When downloading trial information, the user can specify an annotation to all records that are downloaded. By default, annotations are accumulated if trial records are loaded again or updated; alternatively, annotations can be replaced.

Annotations are useful for analyses, for example to specially identify subsets of records in the database.

```{r}
ctrLoadQueryIntoDb(
  queryterm = "https://classic.clinicaltrials.gov/ct2/results?cond=neuroblastoma&recrs=e&age=0&intr=Drug&cntry=DE",
  annotation.text = "site_DE ",
  annotation.mode = "append",
  con = db
)
# * Found search query from CTGOV: cond=neuroblastoma&recrs=e&age=0&intr=Drug&cntry=DE
# Checking helper binaries: done
# (1/3) Checking trials in CTGOV:
# Retrieved overview, records of 12 trial(s) are to be downloaded (estimate: 0.096 MB)
# Download status: 1 done; 0 in progress. Total size: 70.25 Kb (100%)... done!             
# (2/3) Converting to JSON, 12 records converted
# (3/3) Importing JSON records into database...
# = Imported or updated 12 trial(s)                
# = Annotated retrieved records (12 records)
# Updated history ("meta-info" in "test")
```

## Find synonyms of active substance names

Not all registers automatically expand search terms to include alternative terms, such as codes and other names of active substances. To obtain a character vector of synonyms for any active substance name, use:

```{r}
ctrFindActiveSubstanceSynonyms(
  activesubstance = "imatinib"
)
# [1] "imatinib"  "gleevec" "sti 571" "glivec" "CGP 57148" "st1571"
```

These names can then be used in queries in any register.

```{r, include=FALSE}
# cleanup
unlink("sqlite_file.sql")
```

## Using a MongoDB database

This example works with a free service [here](https://www.mongodb.com/cloud/atlas). Note that the user name and password need to be encoded. The format of the connection string is documented at https://docs.mongodb.com/manual/reference/connection-string/.

For recommended databases, see vignette [Install R package ctrdata](file:///Users/ralfherold/Daten/mak/r/emea/ctrdata/docs/articles/ctrdata_install.html#databases-to-use).

```{r remote_mongo}

# Specify base uri for remote MongoDB server,
#  as part of the encoded connection string
db <- nodbi::src_mongo(
  # Note: this provides read-only access
  url = "mongodb+srv://DWbJ7Wh:bdTHh5cS@cluster0-b9wpw.mongodb.net",
  db = "dbperm",
  collection = "dbperm")

# Since the above access is read-only,
# just obtain fields of interest:
dbGetFieldsIntoDf(
  fields = c("a2_eudract_number",
             "e71_human_pharmacology_phase_i"),
  con = db)
#                  _id a2_eudract_number e71_human_pharmacology_phase_i
# 1 2010-024264-18-3RD    2010-024264-18                           TRUE
# 2  2010-024264-18-AT    2010-024264-18                           TRUE
# 3  2010-024264-18-DE    2010-024264-18                           TRUE
# 4  2010-024264-18-GB    2010-024264-18                           TRUE
# 5  2010-024264-18-IT    2010-024264-18                           TRUE
# 6  2010-024264-18-NL    2010-024264-18                           TRUE
```
