---
title: "Retrieve clinical trial information"
author: "Ralf Herold"
date: 2025-06-15
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Retrieve clinical trial information}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Get started

### Attach package `ctrdata`

```{r}
library(ctrdata)
citation("ctrdata")
```

Remember to respect the registers' terms and conditions (see `ctrOpenSearchPagesInBrowser(copyright = TRUE)`). Please cite this package in any publication as follows: Ralf Herold (2025). ctrdata: Retrieve and Analyze Clinical Trials in Public Registers. R package version 1.22.3. https://cran.r-project.org/package=ctrdata

### Open register's advanced search page in browser

These functions open the browser, where the user can start searching for trials of interest.

```{r}
# Please review and respect register copyrights:
ctrOpenSearchPagesInBrowser(
  copyright = TRUE
)

# Open browser with example search:
ctrOpenSearchPagesInBrowser(
  url = "cancer&age=under-18",
  register = "EUCTR"
)
```

### Adjust search parameters and execute search in browser

Refine the search until the trials of interest are listed in the browser. The total number of trials that can be retrieved with package `ctrdata` is intentionally limited to queries with at most 10,000 result records.

### Copy address from browser address bar to clipboard

Use functions or keyboard shortcuts according to the operating system. 

See [here](https://rfhb.github.io/ctrdata/#id_2-script-to-automatically-copy-users-query-from-web-browser) for our automation to copy the URLs of a user's queries in any of the supported clinical trial registers. 

### Get address from clipboard

The next steps are executed in the R environment:

```{r, include=FALSE}
q <- "https://www.clinicaltrialsregister.eu/ctr-search/search?query=cancer&age=under-18&resultsstatus=trials-with-results"
clipr::write_clip(q)
```

```{r}
q <- ctrGetQueryUrl()
# * Using clipboard content as register query URL: https://www.clinicaltrialsregister.eu/
# ctr-search/search?query=cancer&age=under-18&resultsstatus=trials-with-results
# * Found search query from EUCTR: query=cancer&age=under-18&resultsstatus=trials-with-results

q
#                                                    query-term  query-register
# 1 query=cancer&age=under-18&resultsstatus=trials-with-results           EUCTR

# To check, this opens a browser with the query
ctrOpenSearchPagesInBrowser(url = q)
```

### Connect database, retrieve, transform, save and check

Note that in addition to protocol- and results-related information, also all trial documents that made publicly available by the registers, including any protocols, consent forms and results reports, can be downloaded by `ctrdata`, by specifying parameter `documents.path`, see `help(ctrLoadQueryIntoDb)`.

```{r}
# Count number of trial records
ctrLoadQueryIntoDb(
  queryterm = q,
  only.count = TRUE
)$n
# * Checking trials in EUCTR...
# Retrieved overview, multiple records of 386 trial(s) from 20 page(s) to be 
# downloaded (estimate: 50 MB)
# [1] 386

# Connect to a database and chose a collection (table)
db <- nodbi::src_sqlite(
  dbname = "database_name.sql",
  collection = "test"
)

# Retrieve records, download into database
ctrLoadQueryIntoDb(
  queryterm = q,
  con = db
)

# Show which queries have been downloaded into database
dbQueryHistory(con = db)
#       query-timestamp query-register query-records
# 1 2025-06-15 22:06:10          EUCTR          1531
#                                                    query-term
# 1 query=cancer&age=under-18&resultsstatus=trials-with-results
```

With any database, this takes about 80 seconds for 1531 records of 386 trials (~55 ms/record), with most of the time needed for internet-retrieval which is slow from this register. 


## Repeat and update a previous query

Previously executed queries can be repeated by specifying "last" or an integer number for parameter `querytoupdate`, where the number corresponds to the row number of the query shown with `dbQueryHistory()`. Where possible, the query to update first checks for new records in the register. Depending on the register and time since running the query last, an update (differential update) is possible or the original query is executed fully again.

```{r}
# Show all queries
dbQueryHistory(con = db)

# Repeat last query
ctrLoadQueryIntoDb(
  querytoupdate = "last",
  only.count = TRUE,
  con = db
)
# First result page empty - no (new) trials found?
# Updated history ("meta-info" in "test")
```


## Results-related trial information

For CTGOV and CTGOV2, any results are always included in the retrieval. Only for EUCTR, result-related trial information has to be requested to be retrieved, because it will take longer to download and store. No results in structured electronic format are foreseeably available from ISRCTN and CTIS, thus `ctrdata` cannot load them, see `help(ctrLoadQueryIntoDb)`. The download or presence of results is not recorded in `dbQueryHistory()` because the availability of results increases over time. The following only takes about 100 seconds for about 1530 records (~ 65 ms/record) of 386 trials, if run in the same R session as the above commands and thus re-using previous downloads of unprocessed source data. 

```{r}
ctrLoadQueryIntoDb(
  querytoupdate = "last",
  forcetoupdate = TRUE,
  euctrresults = TRUE,
  con = db
)
# * Checking trials in EUCTR...
# Retrieved overview, multiple records of 386 trial(s) from 20 page(s) to be 
# downloaded (estimate: 50 MB)
# (1/3) Downloading trials...
# Note: register server cannot compress data, transfer takes longer (estimate: 500 s)
# (2/3) Converting to NDJSON (estimate: 8 s)...
# (3/3) Importing records into database...
# = Imported or updated 1531 records on 386 trial(s) 
# * Checking results if available from EUCTR for 386 trials: 
# (1/4) Downloading results...
# - extracting results (. = data, F = file[s] and data, x = none):
# . F . . . . . F . F . . . F . F . . F . . . . F . . . . . . . . . . . . . . .
# F . . F . . . . . . . . . . . . . . . . . F . . . . . . . . . . . F . . . . . 
# . . . . . F . . . . . F . . . . . F . . . . . . . . . . . . . . F F . F . . . 
# F . . . . . . . . . F F . F F . . . . F F . . . F . . . . . . . . . . . F . . 
# . . . . . F . . F . . . . F . . . . F . . . . F . . . . . . F F . . . F . . F
# . . . F . . F F F . . . F . . . F F . . . . . . F . F . . . . . . F F F . . F 
# . . . . F F . F . . . . F . . . . . . . . . F . F . . . . . . . . F F . . . . 
# . . . . . . . F . F . . . . . . . . . F . . F F F . . . . F F . . . . F . . .
# F F . . F . . . . . . . . . . F . . . . . . . . . . . F . F . . . . . F . F . 
# . . F . . . . . . . . . . . F F . . F . . . . F F F . . . . . . . . . 
# (2/4) Converting to NDJSON (estimate: 40 s)...
# (3/4) Importing results into database (may take some time)...
# (4/4) Results history: not retrieved (euctrresultshistory = FALSE)
# = Imported or updated results for 386 trials
# Updated history ("meta-info" in "test")
# $n
# [1] 1531
```


## Add trial information from several registers

The same collection can be used to store (and analyse) trial information from different registers, thus can include different and complementary sets of trials. The registers currently supported include CTIS, EUCTR, CTGOV, CTGOV2 and ISRCTN. This can be achieved by loading queries that the user defines specifically or that function `ctrGenerateQueries()` provides, as follows: 

```{r}
# Loading specific query into same collection
ctrLoadQueryIntoDb(
  queryterm = "cond=neuroblastoma&aggFilters=phase:2,ages:child,status:com",
  register = "CTGOV2",
  con = db
)

# Use same query details to obtain queries
queries <- ctrGenerateQueries(
  condition = "neuroblastoma",
  recruitment = "completed",
  phase = "phase 2",
  population = "P"
)

# Open queries in registers' web interfaces
sapply(queries, ctrOpenSearchPagesInBrowser)

# Load all queries into database collection
result <- lapply(queries, ctrLoadQueryIntoDb, con = db)

# Show results of loading
sapply(result, "[[", "n")
# EUCTR       ISRCTN       CTGOV2 CTGOV2expert         CTIS 
#   180            0          110          110            1 

# Overview of queries
dbQueryHistory(con = db)
#       query-timestamp query-register query-records
# 1 2025-06-15 22:06:10          EUCTR          1531
# 2 2025-06-15 22:09:10          EUCTR             0
# 3 2025-06-15 22:21:08         CTGOV2           110
# 4 2025-06-15 22:22:21          EUCTR           180
# 5 2025-06-15 22:22:23         CTGOV2           110
# 6 2025-06-15 22:22:24         CTGOV2           110
# 7 2025-06-15 22:22:26           CTIS             1
# #vquery-term
# 1 query=cancer&age=under-18&resultsstatus=trials-with-results
# 2vquery=cancer&age=under-18&resultsstatus=trials-with-results
# 3vcond=neuroblastoma&aggFilters=phase:2,ages:child,status:com
# 4 query=neuroblastoma&phase=phase-two&age=children&age=adolescent&age=infant-and-toddler&age=newborn&age=preterm-new-born-infants&age=under-18&status=completed
# 5 cond=neuroblastoma&aggFilters=phase:2,ages:child,status:com,studyType:int
# 6 term=AREA[ConditionSearch]"neuroblastoma" AND (AREA[Phase]"PHASE2") AND (AREA[StdAge]"CHILD") AND (AREA[OverallStatus]"COMPLETED") AND (AREA[StudyType]INTERVENTIONAL)
# 7 searchCriteria={"medicalCondition":"neuroblastoma","trialPhaseCode":[4],"ageGroupCode":[2],"status":[5,8]}
```


## Add personal annotations

When loading trial information, the user can specify an annotation string to each of the records that are loaded when calling `ctrLoadQueryIntoDb()`. 
By default, new annotations are appended to any existing annotation of the trial record; alternatively, annotations can be replaced. Annotations are useful for analyses, for example to specially identify subsets of records and trials of interest in the collection

```{r}
# Annotate a query in CTGOV2 defined above
ctrLoadQueryIntoDb(
  queryterm = queries["CTGOV2"],
  annotation.text = "site_DE ",
  annotation.mode = "append",
  con = db
)
# * Appears specific for CTGOV REST API 2.0
# * Found search query from CTGOV2: cond=neuroblastoma&aggFilters=phase:2,ages:child,status:com
# * Checking trials using CTGOV REST API 2.0, found 111 trials
# (1/3) Downloading in 1 batch(es) (max. 1000 trials each; estimate: 11 Mb total)
# (2/3) Converting to NDJSON...
# (3/3) Importing records into database...
# JSON file #: 1 / 1                               
# = Imported or updated 111 trial(s)
# = Annotated retrieved records (111 records)
# Updated history ("meta-info" in "test")
# $n
# [1] 111
```


## Find synonyms of active substance names

Not all registers automatically expand search terms to include alternative terms, such as codes and other names of active substances. The synonymous names can be used in queries in a register that does not offer search expansion. To obtain a character vector of synonyms for an active substance name: 

```{r}
# Search for synonyms
ctrFindActiveSubstanceSynonyms(
  activesubstance = "imatinib"
)
#  [1] "imatinib"          "CGP 57148"         "CGP 57148B"       
#  [4] "CGP57148B"         "Gleevec"           "GLIVEC"           
#  [7] "Imatinib"          "Imatinib Mesylate" "NSC 716051"       
# [10] "ST1571"            "STI 571"           "STI571"   
```

```{r, include=FALSE}
# cleanup
unlink("database_name.sql")
```


## Find platform trials

The interest is increasing to design and use integrated research platforms, clinical research platforms, platform trials, multi-arm multi-stage (MAMS) and master protocol research programs (MPRPs). Additional concepts and terms used include basket and umbrella trials, and in particular complex trials. Please see the references below for further information. `ctrdata` can help finding such research and analysing the study information, as follows: 

```{r}
# Generate queries to identify trials
queries <- ctrGenerateQueries(
  searchPhrase = paste0(
    "basket OR platform OR umbrella OR master protocol OR ",
    "multiarm OR multistage OR subprotocol OR substudy OR ",
    "multi-arm OR multi-stage OR sub-protocol OR sub-study"), 
  startAfter = "2015-01-01")

# See 
help("ctrGenerateQueries")

# Open queries in register web interface
sapply(queries, ctrOpenSearchPagesInBrowser)

# Count number of studies found in the register
result <- lapply(queries, ctrLoadQueryIntoDb, only.count = TRUE)

sapply(result, "[[", "n")
# EUCTR       ISRCTN       CTGOV2 CTGOV2expert         CTIS 
#  1635          165         1538         1538          231

# and see examples in
vignette("ctrdata_summarise")

# Load studies, include EUCTR results data for analysis
result <- lapply(queries, ctrLoadQueryIntoDb, con = db, euctrresults = TRUE)

# See next section for adding related trials
```

References:

- [EU-PEARL. 2023. ‘D2.1. Report on Terminology, References and Scenarios for Platform Trials and Master Protocols’](https://web.archive.org/web/20230314024441/https://eu-pearl.eu/wp-content/uploads/2020/06/EU-PEARL_D2.1_Report-on-Terminology-and-Scenarios-for-Platform-Trials-and-Masterprotocols.pdf)
- [EC / EMA / HMA. 2022. ‘Complex Clinical Trials – Questions and Answers’. EMA/298712/2022](https://health.ec.europa.eu/system/files/2022-06/medicinal_qa_complex_clinical-trials_en.pdf)
- [Williams RJ et al. 2022. ‘Approach for Reporting Master Protocol Study Designs on ClinicalTrials.Gov: Qualitative Analysis’. The BMJ 377 (June):e067745](https://doi.org/10.1136/bmj-2021-067745)
- [Clinical Trials Facilitation and Coordination Group (CTFG). 2019. ‘Recommendation Paper on the Initiation and Conduct of Complex Clinical Trials’](https://www.hma.eu/fileadmin/dateien/Human_Medicines/01-About_HMA/Working_Groups/CTFG/2019_02_CTFG_Recommendation_paper_on_Complex_Clinical_Trials.pdf)


## Load information using trial identifiers

When identifiers of clinical trials of interest are already known, this example shows how they can be processed to import the trial information into a database collection. This involves constructing a query that combines the identifiers and then iterating over the sets of identifiers. Note to combine identifiers into the `queryterm` depends on the specific register. 

```{r}
# Use a trial concept to calculate related identifiers
help("ctrdata-trial-concepts")

# Get data from trials loaded above
df <- dbGetFieldsIntoDf(
  fields = "ctrname",
  calculate = c(
    "f.isUniqueTrial",
    "f.likelyPlatformTrial",
    "f.trialTitle"
  ),
  con = db
)

# Show names of calculated columns in the
# data frame with possible platform trials
names(df)
# [1] "_id"                 
# [2] "ctrname"             
# [3] ".isUniqueTrial"      
# [4] ".likelyPlatformTrial"
# [5] ".likelyRelatedTrials"
# [6] ".maybeRelatedTrials" 
# [7] ".trialTitle" 

# Reduce to unique trials
df <- df[df$.isUniqueTrial, ]
nrow(df)

# Number of recognised set of trials
length(unique(df$.maybeRelatedTrials))
# 571

# Trials with which _id are missing? 
missingIds <- na.omit(setdiff(unlist(df$.maybeRelatedTrials), df$`_id`))

# Load missing trials by _id
res <- list()
for (i in seq_along(missingIds)) {
  message(i, ": ", missingIds[i])
  res <- c(res, suppressMessages(
    list(ctrLoadQueryIntoDb(missingIds[i], euctrresults = TRUE, con = db))))
}

# Trials that could not be loaded are likely phase 1 trials 
# which are not publicly accessible in the in EUCTR register
missingIds[which(sapply(res, "[[", "n") == 0L)]
```

The above loads one trial after the other, just using the `_id` of the trial, from which `ctrdata` infers the concerned register. Alternatively, batches of `_id`s can be loaded from some registers (not CTIS), as follows. 

```{r}
# ids of trials of interest
ctIds <- c(
  "NCT00001209", "NCT00001436", "NCT00187109", "NCT01516567", "NCT01471782", 
  "NCT00357084", "NCT00357500", "NCT00365755", "NCT00407433", "NCT00410657", 
  "NCT00436852", "NCT00445965", "NCT00450307", "NCT00450827", "NCT00471679", 
  "NCT00492167", "NCT00499616", "NCT00503724")

# split into sets of each 10 trial ids 
# (larger sets e.g. 50 may still work)
idSets <- split(ctIds, ceiling(seq_along(ctIds) / 10))

# variable to collect import results
result <- NULL

# iterate over sets of trial ids
for (idSet in idSets) {
  
  setResult <- ctrLoadQueryIntoDb(
    queryterm = paste0("term=", paste0(idSet, collapse = " ")),
    register = "CTGOV2",
    con = db
  )
  
  # check that queried ids have
  # successfully been loaded
  stopifnot(identical(
    sort(setResult$success), sort(idSet)))
  
  # append result
  result <- c(result, list(setResult))
}

# inspect results 
as.data.frame(do.call(rbind, result))[, c("n", "failed")]
#    n failed
# 1 10   NULL
# 2  8   NULL

# queryterms for other registers for retrieving trials by their identifier: 
#
# CTIS (note the comma separated values): 
# https://euclinicaltrials.eu/ctis-public/search#searchCriteria=
# {"containAny":"2025-521008-22-00, 2024-519446-67-00, 2024-517647-31-00"}
#
# EUCTR (note the country suffix os to be removed, values separated with OR): 
# https://www.clinicaltrialsregister.eu/ctr-search/search?
# query=2008-001606-16+OR+2008-001721-34+OR+2008-002260-33
#
```


```{r, include=FALSE}
# cleanup
unlink("database_name.sql")
```


## Example using MongoDB

This example works with a free service [here](https://www.mongodb.com/cloud/atlas). Note that the user name and password need to be encoded. The format of the connection string is documented at https://www.mongodb.com/docs/manual/reference/connection-string/. For recommended databases, see vignette [Install R package ctrdata](https://rfhb.github.io/ctrdata/#databases-that-can-be-used-with-ctrdata).

```{r remote_mongo}
# Specify base uri for remote MongoDB server,
#  as part of the encoded connection string
db <- nodbi::src_mongo(
  # Note: this provides read-only access
  url = "mongodb+srv://DWbJ7Wh:bdTHh5cS@cluster0-b9wpw.mongodb.net",
  db = "dbperm",
  collection = "dbperm")

# Since the above access is read-only,
# just obtain fields of interest:
dbGetFieldsIntoDf(
  fields = c("a2_eudract_number",
             "e71_human_pharmacology_phase_i"),
  con = db)
#                  _id a2_eudract_number e71_human_pharmacology_phase_i
# 1 2010-024264-18-3RD    2010-024264-18                           TRUE
# 2  2010-024264-18-AT    2010-024264-18                           TRUE
# 3  2010-024264-18-DE    2010-024264-18                           TRUE
# 4  2010-024264-18-GB    2010-024264-18                           TRUE
# 5  2010-024264-18-IT    2010-024264-18                           TRUE
# 6  2010-024264-18-NL    2010-024264-18                           TRUE
```
